New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-19 paper/paper.tex
result sections for maximum likelihood approach

A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities. This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$. The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$, if $\symRegRange \geq 30$. For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.

@@

\subsection{Performance}
In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch. For different cardinalities we loaded the precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory. The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm. For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}. Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on-demand or taken from a lookup table. All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.

@@

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
As in \cref{sec:corrected_raw_estimation_error}, where we have validated the corrected raw estimation algorithm, we ran again tests using the maximum likelihood estimation algorithm for different HyperLogLog parameters.

see \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52},\cref{fig:estimation_error_12_14}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}

@@

\subsection{Performance}
see \cref{fig:avg_exec_time}
%In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch. For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory. The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the new estimation algorithm. For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz. As can be seen, the average calculation time did never exceed 700\,ns.

@@

For $\symX \geq 0$ $\symHelper'(\symX)\in[0,0.5]$. Hence, there exists a $\symBeta\in[0,0.5]$ for which

>>>>

A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities. This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$. The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$. For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.

@@

\subsection{Performance}
\label{sec:corrected_raw_estimation_algorithm}
In order to evaluate the performance of the corrected raw estimation algorithm, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch. For different cardinalities we loaded the precalculated multiplicity vectors of 1000 randomly generated HyperLogLog sketches into main memory. The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the algorithm. For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:corrected_raw_avg_exec_time}. Two variants of \cref{alg:corrected_raw_estimation} have been evaluated for which the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ have been either calculated on-demand or taken from a lookup table. All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz.

@@

\subsection{Estimation error}
\label{sec:maximum_likelihood_estimation_error}
In order to investigate the estimation error for the maximum likelihood estimation algorithm, we investigated the same HyperLogLog configurations as in \cref{sec:corrected_raw_estimation_error} for the corrected raw estimation algorithm. \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52}, and \cref{fig:estimation_error_12_14} show very similar results for various HyperLogLog parameters.

What is different for the maximum likelihood estimation approach is a somewhat smaller median bias with less oscillations around zero for small cardinalities. Furthermore, contrary to the raw estimator which reveals a small oscillating bias for the mean (see \cref{fig:raw_corrected_estimation_error_22_10}), the maximum likelihood estimator seems to be completely unbiased (see \cref{fig:estimation_error_22_10}).

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}

@@

\subsection{Performance}
We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. The average computation time shows a different behavior over the cardinality than those of the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}). As can be seen, the average calculation time is longer for most cardinalities, but nevertheless it never exceeds 700\,ns which is still fast enough for many applications.

@@

Since $\symHelper'(\symX)\in[0,0.5]$ for $\symX \geq 0$, there exists a $\symBeta\in[0,0.5]$ for which
